---
title: "Logistics Delivery Analytics Platform"
subtitle: "End-to-End ETL Pipeline & Operational Analytics Report"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5
)

required_packages <- c(
  "tidyverse",
  "knitr",
  "scales"
)

missing_packages <- required_packages[
  !required_packages %in% installed.packages()[, "Package"]
]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

library(tidyverse)
library(knitr)
library(scales)
```

```{r load-data}
delivery <- read_csv(
  "fact_delivery_performance.csv",
  show_col_types = FALSE
)

delivery <- delivery %>%
  mutate(
    order_date = as.Date(order_date),
    delivery_date = as.Date(delivery_date),
    transit_days = as.numeric(delivery_date - order_date),
    partner = as.factor(partner),
    province = as.factor(province),
    weather = as.factor(weather),
    traffic_level = factor(
      traffic_level,
      levels = c("Low", "Medium", "High")
    )
  )
```

# Executive Summary

This project analyzes delivery performance data to identify the operational factors associated with delivery speed, cost, and efficiency.

The final dataset was produced through an end-to-end data pipeline involving Azure Blob Storage and MS SQL Server.

The analysis evaluates:

* Delivery partner performance
* Delivery costs and cost efficiency
* Weather and traffic conditions
* Provincial and city-level performance
* Distance and delivery-time relationships
* Average delivery speed

```{r executive-kpis}
overall_kpis <- tibble(
  Metric = c(
    "Total Deliveries",
    "Average Delivery Time",
    "Average Delivery Cost",
    "Average Distance",
    "Average Cost per Kilometre",
    "Average Delivery Speed"
  ),
  Value = c(
    comma(nrow(delivery)),
    paste0(round(mean(delivery$delivery_time_hours, na.rm = TRUE), 2), " hours"),
    dollar(mean(delivery$cost, na.rm = TRUE)),
    paste0(round(mean(delivery$distance_km, na.rm = TRUE), 1), " km"),
    dollar(mean(delivery$cost_per_km, na.rm = TRUE)),
    paste0(round(mean(delivery$avg_speed_kmh, na.rm = TRUE), 1), " km/h")
  )
)

kable(
  overall_kpis,
  caption = "Overall Delivery Performance Indicators",
  align = c("l", "r")
)
```

# Project Architecture

The project follows this workflow:

**Raw Delivery Data → Azure Blob Storage → SQL ETL → Reporting Fact Table → R Markdown Analysis**

Azure SQL was used to transform the data, remove invalid records, enrich transactions with weather and traffic conditions, and calculate reporting metrics.

The final `fact_delivery_performance` table was exported as a CSV and analyzed in R.

# Data Overview

```{r data-overview}
data_overview <- tibble(
  Measurement = c(
    "Rows",
    "Columns",
    "Unique Delivery Partners",
    "Unique Cities",
    "Unique Provinces",
    "Earliest Order Date",
    "Latest Delivery Date"
  ),
  Result = c(
    nrow(delivery),
    ncol(delivery),
    n_distinct(delivery$partner),
    n_distinct(delivery$city),
    n_distinct(delivery$province),
    format(min(delivery$order_date, na.rm = TRUE), "%B %d, %Y"),
    format(max(delivery$delivery_date, na.rm = TRUE), "%B %d, %Y")
  )
)

kable(
  data_overview,
  caption = "Dataset Overview",
  align = c("l", "r")
)
```

```{r sample-data}
delivery %>%
  select(
    delivery_id,
    order_date,
    delivery_date,
    partner,
    city,
    province,
    weather,
    traffic_level,
    distance_km,
    delivery_time_hours,
    cost
  ) %>%
  head(10) %>%
  kable(
    caption = "Sample of Final Reporting Data",
    digits = 2
  )
```

# Data Quality Validation

```{r data-quality}
quality_checks <- tibble(
  Check = c(
    "Missing Delivery IDs",
    "Duplicate Delivery IDs",
    "Missing Partner Values",
    "Missing Delivery Times",
    "Missing Costs",
    "Missing Weather Values",
    "Missing Traffic Values",
    "Invalid Distances",
    "Invalid Costs"
  ),
  Records = c(
    sum(is.na(delivery$delivery_id)),
    sum(duplicated(delivery$delivery_id)),
    sum(is.na(delivery$partner)),
    sum(is.na(delivery$delivery_time_hours)),
    sum(is.na(delivery$cost)),
    sum(is.na(delivery$weather)),
    sum(is.na(delivery$traffic_level)),
    sum(delivery$distance_km <= 0, na.rm = TRUE),
    sum(delivery$cost <= 0, na.rm = TRUE)
  )
)

kable(
  quality_checks,
  caption = "Final Dataset Quality Checks",
  align = c("l", "r")
)
```

The validation results confirm whether the final reporting dataset contains duplicate identifiers, missing business-critical fields, or invalid numerical values.

# Delivery Partner Performance

```{r partner-summary}
partner_summary <- delivery %>%
  group_by(partner) %>%
  summarise(
    deliveries = n(),
    average_delivery_time = mean(delivery_time_hours, na.rm = TRUE),
    average_cost = mean(cost, na.rm = TRUE),
    average_cost_per_km = mean(cost_per_km, na.rm = TRUE),
    average_speed = mean(avg_speed_kmh, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(average_delivery_time)

partner_summary %>%
  mutate(
    average_delivery_time = round(average_delivery_time, 2),
    average_cost = dollar(average_cost),
    average_cost_per_km = dollar(average_cost_per_km),
    average_speed = round(average_speed, 2)
  ) %>%
  rename(
    Partner = partner,
    Deliveries = deliveries,
    `Average Delivery Time (Hours)` = average_delivery_time,
    `Average Cost` = average_cost,
    `Average Cost per KM` = average_cost_per_km,
    `Average Speed (KM/H)` = average_speed
  ) %>%
  kable(
    caption = "Delivery Partner Performance Summary",
    align = c("l", "r", "r", "r", "r", "r")
  )
```

## Average Delivery Time by Partner

```{r partner-time-chart}
partner_summary %>%
  ggplot(
    aes(
      x = reorder(partner, average_delivery_time),
      y = average_delivery_time
    )
  ) +
  geom_col() +
  coord_flip() +
  geom_text(
    aes(label = round(average_delivery_time, 2)),
    hjust = -0.1,
    size = 3.5
  ) +
  labs(
    title = "Average Delivery Time by Partner",
    subtitle = "Lower delivery times indicate faster service",
    x = NULL,
    y = "Average Delivery Time (Hours)"
  ) +
  expand_limits(
    y = max(partner_summary$average_delivery_time) * 1.15
  ) +
  theme_minimal()
```

```{r fastest-partner-text, results='asis'}
fastest_partner <- partner_summary %>%
  slice_min(average_delivery_time, n = 1, with_ties = FALSE)

cat(
  "**Finding:**",
  as.character(fastest_partner$partner),
  "had the lowest average delivery time at",
  round(fastest_partner$average_delivery_time, 2),
  "hours across",
  fastest_partner$deliveries,
  "deliveries."
)
```

## Average Delivery Cost by Partner

```{r partner-cost-chart}
partner_summary %>%
  ggplot(
    aes(
      x = reorder(partner, average_cost),
      y = average_cost
    )
  ) +
  geom_col() +
  coord_flip() +
  geom_text(
    aes(label = dollar(average_cost)),
    hjust = -0.1,
    size = 3.5
  ) +
  labs(
    title = "Average Delivery Cost by Partner",
    subtitle = "Comparison of average transaction cost",
    x = NULL,
    y = "Average Delivery Cost"
  ) +
  scale_y_continuous(labels = dollar) +
  expand_limits(
    y = max(partner_summary$average_cost) * 1.18
  ) +
  theme_minimal()
```

## Cost Efficiency by Partner

```{r partner-efficiency-chart}
partner_summary %>%
  ggplot(
    aes(
      x = reorder(partner, average_cost_per_km),
      y = average_cost_per_km
    )
  ) +
  geom_col() +
  coord_flip() +
  geom_text(
    aes(label = dollar(average_cost_per_km)),
    hjust = -0.1,
    size = 3.5
  ) +
  labs(
    title = "Average Cost per Kilometre by Partner",
    subtitle = "Lower cost per kilometre indicates stronger cost efficiency",
    x = NULL,
    y = "Average Cost per Kilometre"
  ) +
  scale_y_continuous(labels = dollar) +
  expand_limits(
    y = max(partner_summary$average_cost_per_km) * 1.18
  ) +
  theme_minimal()
```

```{r efficient-partner-text, results='asis'}
efficient_partner <- partner_summary %>%
  slice_min(average_cost_per_km, n = 1, with_ties = FALSE)

cat(
  "**Finding:**",
  as.character(efficient_partner$partner),
  "had the lowest average cost per kilometre at",
  dollar(efficient_partner$average_cost_per_km),
  "per kilometre."
)
```

# Weather Impact

```{r weather-summary}
weather_summary <- delivery %>%
  group_by(weather) %>%
  summarise(
    deliveries = n(),
    average_delivery_time = mean(delivery_time_hours, na.rm = TRUE),
    average_speed = mean(avg_speed_kmh, na.rm = TRUE),
    average_cost = mean(cost, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(average_delivery_time)

weather_summary %>%
  mutate(
    average_delivery_time = round(average_delivery_time, 2),
    average_speed = round(average_speed, 2),
    average_cost = dollar(average_cost)
  ) %>%
  rename(
    Weather = weather,
    Deliveries = deliveries,
    `Average Delivery Time (Hours)` = average_delivery_time,
    `Average Speed (KM/H)` = average_speed,
    `Average Cost` = average_cost
  ) %>%
  kable(
    caption = "Performance by Weather Condition"
  )
```

```{r weather-chart}
weather_summary %>%
  ggplot(
    aes(
      x = reorder(weather, average_delivery_time),
      y = average_delivery_time
    )
  ) +
  geom_col() +
  geom_text(
    aes(label = round(average_delivery_time, 2)),
    vjust = -0.4
  ) +
  labs(
    title = "Average Delivery Time by Weather Condition",
    subtitle = "Comparison of delivery speed across weather categories",
    x = "Weather Condition",
    y = "Average Delivery Time (Hours)"
  ) +
  expand_limits(
    y = max(weather_summary$average_delivery_time) * 1.15
  ) +
  theme_minimal()
```

# Traffic Impact

```{r traffic-summary}
traffic_summary <- delivery %>%
  group_by(traffic_level) %>%
  summarise(
    deliveries = n(),
    average_delivery_time = mean(delivery_time_hours, na.rm = TRUE),
    average_speed = mean(avg_speed_kmh, na.rm = TRUE),
    average_cost = mean(cost, na.rm = TRUE),
    .groups = "drop"
  )

traffic_summary %>%
  mutate(
    average_delivery_time = round(average_delivery_time, 2),
    average_speed = round(average_speed, 2),
    average_cost = dollar(average_cost)
  ) %>%
  rename(
    `Traffic Level` = traffic_level,
    Deliveries = deliveries,
    `Average Delivery Time (Hours)` = average_delivery_time,
    `Average Speed (KM/H)` = average_speed,
    `Average Cost` = average_cost
  ) %>%
  kable(
    caption = "Performance by Traffic Level"
  )
```

```{r traffic-chart}
traffic_summary %>%
  ggplot(
    aes(
      x = traffic_level,
      y = average_delivery_time
    )
  ) +
  geom_col() +
  geom_text(
    aes(label = round(average_delivery_time, 2)),
    vjust = -0.4
  ) +
  labs(
    title = "Average Delivery Time by Traffic Level",
    subtitle = "Comparison of delivery performance under different traffic conditions",
    x = "Traffic Level",
    y = "Average Delivery Time (Hours)"
  ) +
  expand_limits(
    y = max(traffic_summary$average_delivery_time) * 1.15
  ) +
  theme_minimal()
```

# Provincial Performance

```{r province-summary}
province_summary <- delivery %>%
  group_by(province) %>%
  summarise(
    deliveries = n(),
    average_delivery_time = mean(delivery_time_hours, na.rm = TRUE),
    average_cost = mean(cost, na.rm = TRUE),
    average_distance = mean(distance_km, na.rm = TRUE),
    average_speed = mean(avg_speed_kmh, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(average_delivery_time)

province_summary %>%
  mutate(
    average_delivery_time = round(average_delivery_time, 2),
    average_cost = dollar(average_cost),
    average_distance = round(average_distance, 2),
    average_speed = round(average_speed, 2)
  ) %>%
  rename(
    Province = province,
    Deliveries = deliveries,
    `Average Delivery Time (Hours)` = average_delivery_time,
    `Average Cost` = average_cost,
    `Average Distance (KM)` = average_distance,
    `Average Speed (KM/H)` = average_speed
  ) %>%
  kable(
    caption = "Provincial Delivery Performance"
  )
```

```{r province-chart}
province_summary %>%
  ggplot(
    aes(
      x = reorder(province, average_delivery_time),
      y = average_delivery_time
    )
  ) +
  geom_col() +
  geom_text(
    aes(label = round(average_delivery_time, 2)),
    vjust = -0.4
  ) +
  labs(
    title = "Average Delivery Time by Province",
    subtitle = "Regional comparison of delivery performance",
    x = "Province",
    y = "Average Delivery Time (Hours)"
  ) +
  expand_limits(
    y = max(province_summary$average_delivery_time) * 1.15
  ) +
  theme_minimal()
```

# City-Level Performance

```{r city-summary}
city_summary <- delivery %>%
  group_by(city) %>%
  summarise(
    deliveries = n(),
    average_delivery_time = mean(delivery_time_hours, na.rm = TRUE),
    average_cost = mean(cost, na.rm = TRUE),
    average_speed = mean(avg_speed_kmh, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(average_delivery_time)

city_summary %>%
  mutate(
    average_delivery_time = round(average_delivery_time, 2),
    average_cost = dollar(average_cost),
    average_speed = round(average_speed, 2)
  ) %>%
  rename(
    City = city,
    Deliveries = deliveries,
    `Average Delivery Time (Hours)` = average_delivery_time,
    `Average Cost` = average_cost,
    `Average Speed (KM/H)` = average_speed
  ) %>%
  kable(
    caption = "City-Level Delivery Performance"
  )
```

# Distance and Delivery Time

```{r distance-time-chart}
ggplot(
  delivery,
  aes(
    x = distance_km,
    y = delivery_time_hours
  )
) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  labs(
    title = "Relationship Between Distance and Delivery Time",
    subtitle = "The trend line shows how delivery time changes as distance increases",
    x = "Distance (KM)",
    y = "Delivery Time (Hours)"
  ) +
  theme_minimal()
```

```{r distance-correlation, results='asis'}
distance_time_correlation <- cor(
  delivery$distance_km,
  delivery$delivery_time_hours,
  use = "complete.obs"
)

cat(
  "**Finding:** The correlation between delivery distance and delivery time is",
  round(distance_time_correlation, 2),
  "."
)
```

# Distance and Cost

```{r distance-cost-chart}
ggplot(
  delivery,
  aes(
    x = distance_km,
    y = cost
  )
) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Relationship Between Distance and Delivery Cost",
    subtitle = "The trend line shows how delivery cost changes with distance",
    x = "Distance (KM)",
    y = "Delivery Cost"
  ) +
  theme_minimal()
```

```{r cost-correlation, results='asis'}
distance_cost_correlation <- cor(
  delivery$distance_km,
  delivery$cost,
  use = "complete.obs"
)

cat(
  "**Finding:** The correlation between delivery distance and delivery cost is",
  round(distance_cost_correlation, 2),
  "."
)
```

# Delivery Speed Analysis

```{r speed-chart}
partner_summary %>%
  ggplot(
    aes(
      x = reorder(partner, average_speed),
      y = average_speed
    )
  ) +
  geom_col() +
  coord_flip() +
  geom_text(
    aes(label = round(average_speed, 1)),
    hjust = -0.1,
    size = 3.5
  ) +
  labs(
    title = "Average Delivery Speed by Partner",
    subtitle = "Higher average speed indicates faster movement across delivery distances",
    x = NULL,
    y = "Average Speed (KM/H)"
  ) +
  expand_limits(
    y = max(partner_summary$average_speed) * 1.15
  ) +
  theme_minimal()
```

# Key Findings

```{r key-findings, results='asis'}
slowest_partner <- partner_summary %>%
  slice_max(average_delivery_time, n = 1, with_ties = FALSE)

lowest_cost_partner <- partner_summary %>%
  slice_min(average_cost, n = 1, with_ties = FALSE)

highest_speed_partner <- partner_summary %>%
  slice_max(average_speed, n = 1, with_ties = FALSE)

slowest_weather <- weather_summary %>%
  slice_max(average_delivery_time, n = 1, with_ties = FALSE)

slowest_traffic <- traffic_summary %>%
  filter(!is.na(traffic_level)) %>%
  slice_max(average_delivery_time, n = 1, with_ties = FALSE)

best_province <- province_summary %>%
  slice_min(average_delivery_time, n = 1, with_ties = FALSE)

cat(
  paste0(
    "1. **Fastest delivery partner:** ",
    fastest_partner$partner,
    " recorded the lowest average delivery time of ",
    round(fastest_partner$average_delivery_time, 2),
    " hours.\n\n"
  ),
  paste0(
    "2. **Slowest delivery partner:** ",
    slowest_partner$partner,
    " recorded the highest average delivery time of ",
    round(slowest_partner$average_delivery_time, 2),
    " hours.\n\n"
  ),
  paste0(
    "3. **Lowest average delivery cost:** ",
    lowest_cost_partner$partner,
    " had an average cost of ",
    dollar(lowest_cost_partner$average_cost),
    ".\n\n"
  ),
  paste0(
    "4. **Strongest cost efficiency:** ",
    efficient_partner$partner,
    " recorded the lowest average cost per kilometre at ",
    dollar(efficient_partner$average_cost_per_km),
    ".\n\n"
  ),
  paste0(
    "5. **Highest delivery speed:** ",
    highest_speed_partner$partner,
    " recorded an average speed of ",
    round(highest_speed_partner$average_speed, 2),
    " km/h.\n\n"
  ),
  paste0(
    "6. **Weather condition with the longest delivery time:** ",
    slowest_weather$weather,
    " averaged ",
    round(slowest_weather$average_delivery_time, 2),
    " hours.\n\n"
  ),
  paste0(
    "7. **Traffic condition with the longest delivery time:** ",
    slowest_traffic$traffic_level,
    " traffic averaged ",
    round(slowest_traffic$average_delivery_time, 2),
    " hours.\n\n"
  ),
  paste0(
    "8. **Fastest province:** ",
    best_province$province,
    " recorded the lowest average delivery time of ",
    round(best_province$average_delivery_time, 2),
    " hours."
  )
)
```

# Business Recommendations

1. **Prioritize high-performing partners.** Use delivery time, delivery speed, and cost-per-kilometre metrics together when assigning shipments.

2. **Investigate inefficient partners.** Partners with higher costs and slower delivery times should be reviewed for route, capacity, or service issues.

3. **Include weather and traffic in delivery planning.** Historical condition data can support more realistic delivery estimates and proactive customer communication.

4. **Use distance-based pricing controls.** Cost-per-kilometre should be monitored to identify expensive routes and pricing inconsistencies.

5. **Review regional performance.** Provinces or cities with slower average delivery times may require route redesign, additional carrier capacity, or improved scheduling.

6. **Expand future data collection.** Future versions should include promised delivery dates, delay reasons, package types, customer ratings, and delivery status to support stronger service-level analysis.

# Limitations

The analysis is based on a small synthetic dataset and should be interpreted as a demonstration of data engineering and analytical methods rather than a production logistics assessment.

The dataset does not contain a promised delivery deadline or an explicit on-time indicator. Therefore, the report compares delivery speed and efficiency rather than calculating an official on-time delivery rate.

Weather and traffic records were joined using the closest available condition record within the same province, with preference given to matching cities. This approach reduced missing matches but may not represent the exact conditions at the delivery time.

# Conclusion

This project demonstrates a complete analytics workflow from raw logistics data to a final business intelligence report.

Python was used for initial data cleaning, Azure SQL supported ETL and reporting-table creation, and R was used to evaluate carrier performance, cost efficiency, regional patterns, weather, traffic, distance, and delivery speed.

The resulting workflow provides a reusable foundation for larger logistics datasets and more advanced forecasting or optimization models.



